---
title: "Who Goes Home First? Fair Peel-Off Orders for Anaesthetists"
subtitle: "The Anesthetist's Peel Assignment Problem (APAP). Working draft"
author: "Helga Ingimundardóttir & Tómas Philip Rúnarsson"
date: today
abstract: |
  In a hospital anaesthesia department, the working day does not end at a fixed time. Each
  consultant may leave once the operating rooms they cover have closed, and since case durations
  are uncertain, the department agrees each morning on a *peel-off order*: who is released first,
  second, and so on. Late positions mean long days, so the order has to be shared out fairly. We
  formulate this as the Anesthetist's Peel Assignment Problem (APAP), a weekly mixed-integer
  program that assigns peel-off positions to the consultants without a fixed shift. The
  consultants' points per day worked are kept close to a common target, and the cardiac and
  in-charge roles are spread evenly. We apply the model retrospectively to 285 weeks of real call
  schedules (July 2018 to December 2023, 25 consultants) and report how evenly the resulting
  peel-off burden is shared. Individual weeks are closely balanced, but because each week is
  solved on its own, small persistent differences between consultants are not compensated over
  the long run: fair weeks do not by themselves give long-run fairness.
execute:
  echo: false
---

```{r}
#| label: setup
source("_setup.R")
```

::: {.callout-note}
## Draft status

This is a working draft. Items marked **TODO** need input before submission. Doctors are
identified by their initials, as in the schedules they provided for research. **TODO:** replace
them with pseudonymous codes before publication (`python -m data.apap_export --pseudonymise`).
:::

# Introduction {#sec-intro}

Most shift work has a fixed end. Anaesthesia does not. An anaesthetist's day starts before 7 am
and ends when the last operation in their rooms is finished, and that time is not known in
advance: cases overrun, are added, or are cancelled. In a large department many consultants work
in parallel, and as the afternoon goes on, rooms close one by one and the consultants covering
them are no longer needed.

Someone still has to decide who leaves. If consultants were simply released as their own rooms
closed, the length of each person's day would depend on their room allocation, which is set for
clinical reasons and not for fairness. Instead, the department studied here agrees on a
**peel-off order** each day. The first person on the list is released first, the next person
covers any remaining work, and so on down the list. The last people on the list stay until the
end of the day. We use the department's term *peel-off order* for what the anaesthesia literature
generally calls the order of *end-of-shift relief*.

A good position one day and a bad one the next is normal; nobody expects a single day to be fair.
The question is whether the order is fair **over time**. Before this project, the order was set
by hand, and the consultants had no clear view of how their positions added up. This paper
describes a mathematical model, the *Anesthetist's Peel Assignment Problem* (APAP), that sets the
order so that each consultant's burden per day worked is kept close to a common target.

The work was done in the autumn of 2023, when the department provided its historical call
schedules for research, and was written up as a working paper. It was then put on hold while we
waited for more reliable data from the hospital's records on when the consultants actually left
work. Those data never became available. This draft picks the work up again with the data at
hand: the call schedules, and the model's peel-off orders for them.

The paper makes three contributions:

1. It formulates end-of-day relief as a *repeated ordinal allocation problem*: each day the
   consultants receive distinct positions in an order of release, and fairness is assessed
   across repeated allocations and not within a single day (@sec-problem).
2. It gives a weekly mixed-integer programming (MIP) formulation that assigns the complete
   peel-off order, balancing points per day worked while respecting fixed shifts, requests and
   clinical roles (@sec-model).
3. It applies the model retrospectively to 5½ years of real call schedules and examines whether
   fairness within individual weeks carries over to fairness over longer horizons (@sec-data,
   @sec-results).

**Related work.** Personnel rostering is well studied in healthcare, including nurse and
physician scheduling [@ernst2004staff; @burke2004nurse; @erhard2018physician]. These models
decide who works which shift, duty or workstation. Anaesthesia adds uncertain surgery durations
and an uncertain end of day, and optimisation models have covered duty and workstation rosters
with preferences and fairness [@fugener2015duty], the joint assignment of anaesthetists and
operating rooms [@rath2017integrated; @tsang2025stochastic], and equitable workload under
uncertain demand [@sun2023equitable].

A smaller literature deals directly with end-of-shift relief. @wax2017relief built a decision
support system that gives priority for relief to anaesthetists who have recently worked later
than others, and @bhutiani2018relief used a prioritised relief list to make the relief of
residents more equitable. @dexter2021variability studied the variability in work hours among
anaesthetists who are not on call, and @dexter2022relief showed that relief decisions based on
equity can lower the group's productivity in the late afternoon. These studies establish the
order of relief as an equity problem in anaesthesia.

The closest optimisation work keeps the relief order separate from the operational problem that
follows it. @bansal2026relief schedule anaesthetists to operating rooms under uncertain surgery
durations, deciding relief times and cover, and take the relief order as an input. APAP
addresses the question that comes before: how should that order itself be chosen? It does not
apply a history-based priority rule one relief at a time. It assigns the complete daily order,
jointly over a week, while accounting for the fixed call-related positions, requests, days
worked and clinical roles.

APAP sits at the intersection of two strands of work. Equity theory asks how indivisible,
unequal burdens should be distributed among people with different claims [@young1994equity],
and how fairness trades off against efficiency [@bertsimas2011price]. Repeated-scheduling work
asks how fairness should be assessed across a sequence of periods [@ajtai1998fairness;
@gross2019satisfaction; @hermelin2025fairness]. APAP brings these questions together as a
repeated ordinal allocation problem: a single day cannot give everyone an early position, but
advantages and burdens can be balanced across repeated allocations. Our results also show why
the time horizon matters: weeks that are each fair need not add up to long-run equity when past
burden is not carried from one planning period to the next (@sec-discussion).

# Problem setting {#sec-problem}

## The department

The group is a paediatric anaesthesia department in North America with 22–25 consultants. They
work across three sites:

* an **outpatient surgical clinic**, open during the day only, for minor surgery;
* a **specialty hospital** with 24/7 cover for orthopaedics, rehabilitation, neurosurgery and
  plastic surgery, staffed by 3–4 anaesthetists, one of whom stays overnight;
* the **main children's hospital**, the largest site, which needs 11–13 anaesthetists on an
  average day. Two are always on call there: the *call* anaesthetist stays overnight and the
  *late-call* anaesthetist works into the evening.

About half of the consultants are qualified for **cardiac** anaesthesia, and one of them is always
in either the call or the late-call role. One consultant is the **charge** doctor, who manages the
operating rooms for the day and usually works long hours. Consultants who are not working on a
given day are on leave or have administrative time.

## Peel-off positions and points

Each working day, every consultant in the order gets a position $k = 1, 2, \dots$, where $k = 1$
is released first. The position is also the consultant's **points** for the day: a high number
means a long day. Many positions follow directly from the call schedule:

| Shift | Meaning | Position |
|---|---|---|
| PostCall | worked overnight the night before | first (1) |
| PostHoliday | worked on a holiday the day before | 1–2 |
| PostLate | was late-call the day before | 1–3 |
| PreCall | is on call the next day | 2–4 |
| *Unassigned* | no fixed constraint: set by the model | middle positions |
| OnLate | late-call today | second to last |
| OnCall | on call today | last |

: Shift types and their position in the peel-off order. {#tbl-shifts}

@fig-shift-points shows the positions observed for each shift type. On a typical weekday there
are `r median(per_day$in_order)` consultants in the order, and `r median(per_day$pool)` of them
are unassigned. Choosing an order for those consultants is the actual decision. Days with
administrative duties count as a fixed number of points (8), so that such days are not treated
as free.

```{r}
#| label: fig-shift-points
#| fig-cap: "Peel-off positions held on weekdays, by shift type. Fixed shifts take the first and last positions; the unassigned pool fills the middle."
#| fig-height: 3.2
plot_shift_points(apap$assignments)
```

## What fairness means here {#sec-fairness}

@young1994equity defines an *allocation problem* as one in which "a bundle of resources,
burdens, or costs is temporarily held in common by a group and must be allotted to them
individually". The peel-off order is such a problem. The burden is the end of the working day:
someone has to stay until the last room closes, and the group has to decide who.

**The good is indivisible.** A position cannot be split: exactly one consultant is released
first, and exactly one stays last. Young describes two common ways of making indivisible goods
fair. One is *rotation*, where claimants take turns, so that fairness holds over a sequence of
allocations and not within each one. The other is a *point system* that ranks claimants by
their accumulated claim, like the points used to order early discharge from the US Army after
1945 or to allocate kidneys. APAP combines the two: each position carries points, and the
positions rotate so that points even out over the week.

**The good is ordinal.** The daily decision is not the allocation of one scarce resource. It is
the allocation of an ordered set of outcomes, $1 \prec 2 \prec \dots \prec n$, where every
consultant in the order receives exactly one rank. This is what separates APAP from shift
assignment and from ordinary priority scheduling, and it is why we describe it as a *repeated
ordinal allocation problem*. The points turn the ordinal outcomes into an additive measure of
burden that can be balanced over time.

**Claims differ.** Consultants work different numbers of days: some work part-time, some joined
or left during the period, and leave differs from person to person. Equal total points would
therefore not be fair. Following Aristotle's proportionality principle, the burden should be
proportional to the contribution, here the number of days worked. Fairness is therefore measured
as **points per day worked**, and the aim is to keep this ratio close to a common target $c$ for
everyone.

**Apportionment.** There is a useful analogy with the classical apportionment problem of dividing
parliamentary seats among states in proportion to population [@young1994equity]. A consultant's
days worked play the role of population, and the target $c$ plays the role of the divisor. The
principle behind Webster's and Hill's methods, that representation *per capita* should be as near
uniform as possible, becomes the principle that points *per day worked* should be as near uniform
as possible. The analogy is not exact: seats are interchangeable, whereas peel-off positions are
ordered, and a complete order has to be assigned each day. Two further criteria from
apportionment carry over:

* *Impartiality*: consultants who look the same should be treated the same. In particular,
  being qualified for the cardiac or charge role should not in itself change a consultant's
  points per day worked.
* *Unbiasedness*: over many allocations, the difference between each claimant's average
  allotment and their quota should be close to zero. For APAP this is the long-run question:
  do small weekly differences cancel out over months and years, or do they accumulate?

Envy-freeness, where no one prefers another's allocation, is not a useful criterion for a single
day. If earlier release is preferred, every consultant with a later position would prefer an
earlier one, and only one consultant can hold each position. The relevant question is therefore
not whether one day's order is envy-free, but whether advantages and burdens are balanced across
repeated allocations, which again points to rotation and long-run balance.

@tbl-design answers the design questions that @young1994equity poses for any equity rule.

| Design question | APAP |
|---|---|
| What form does the allocation take? | A daily order of release; each position is indivisible |
| Who is eligible? | Consultants working that day without a fixed call-related position |
| What counts? | Peel-off position as points; administrative days count as 8 points |
| Which principle? | Proportionality: equal points per day worked |
| Which precedents? | Fixed positions after call, before call and on call (@tbl-shifts) |
| How are competing criteria reconciled? | Weighted objective: equity first, then role spread, then charge from the call team |
| What incentives does the rule create? | **TODO:** e.g. requests for particular positions, or for administrative days |

: The design questions of @young1994equity, answered for APAP. {#tbl-design}

# Data {#sec-data}

## Source

The data are the department's quarterly call schedules from July 2018 to December 2023, exported
from the scheduling system as spreadsheets. For each day they give who is on call, who is
late-call, who is at which site, who is on leave and who has administrative time. A data
pipeline (see the appendix) converts them to weekly instances:

1. quarterly spreadsheets are converted to a structured format and split into weeks (Monday to
   Sunday);
2. the fixed shifts of @tbl-shifts are derived from each day and its neighbours (for example,
   the call doctor yesterday is *PostCall* today);
3. staff requests for particular positions are added;
4. the model is solved for each week, and the results are collected in one dataset.

The schedules record who worked which shift, but not the peel-off order that was used on each
day, nor when each consultant left. The results below are therefore a *retrospective
application* of the model to real schedules, not an observed history. Registry data on actual
departure times were requested from the hospital but never became available (@sec-intro).

## Overview

```{r}
#| label: tbl-overview
#| tbl-cap: "Overview of the dataset."
n_weeks <- nrow(apap$weeks)
tibble(
  Item = c("Period", "Weeks solved", "Consultants", "Qualified for cardiac / charge / both / neither",
           "Doctor-days in the order", "of which weekdays"),
  Value = c(paste(format(min(apap$weeks$start), "%B %Y"), "to", format(max(apap$weeks$end), "%B %Y")),
            n_weeks, nrow(apap$doctors),
            paste(table(apap$doctors$roles)[c("Cardiac", "Charge", "Both", "Neither")], collapse = " / "),
            format(nrow(apap$assignments), big.mark = ","), format(nrow(weekday), big.mark = ","))
) |> knitr::kable()
```

@fig-days-worked shows how many days each consultant worked. About half worked between 1,000 and 1,150 days
in the period. The consultants with fewer days worked part-time or were only in the group for part
of the period. On weekends and holidays only the call and late-call consultants work.

```{r}
#| label: fig-days-worked
#| fig-cap: "Days worked per consultant in the period, by day type. Consultants are ordered by average weekly points."
#| fig-height: 3.5
plot_days_worked(apap$assignments, apap$points)
```

# Model {#sec-model}

The model is solved for one week at a time. Within a week, it assigns positions to the unassigned
consultants each day, chooses the cardiac and charge doctors for each weekday, and keeps each
consultant's points per day worked close to a target.

## Sets and parameters

| Symbol | Meaning |
|---|---|
| $D$ | consultants working in the week |
| $T$, $T^w \subseteq T$ | days of the week, and weekdays |
| $W_t \subseteq D$ | consultants in the peel-off order on day $t$ |
| $U_t \subseteq W_t$ | unassigned consultants on day $t$ |
| $K_t = \{1, \dots, \bar k_t\}$ | positions on day $t$ |
| $\pi_t(d)$ | fixed position of consultant $d$ with a fixed shift on day $t$ |
| $L_t$ | call and late-call consultants on day $t$ |
| $A_t$ | consultants with administrative time on day $t$ |
| $C$, $H$ | consultants qualified for cardiac and for charge |
| $\kappa_t$ | position that the charge doctor must hold on day $t$ |
| $n_d$ | weekdays worked by $d$ in the week |
| $a = 8$ | points for an administrative day |
| $E = \{1, 0.5, 0.2\}$ | tolerance bands around the target |

## Decision variables

* $x_{dtk} \in \{0,1\}$: consultant $d$ holds position $k$ on day $t$;
* $c \ge 0$: the week's target, in points per day worked (free, chosen by the model);
* $y^\varepsilon_d \in \{0,1\}$: consultant $d$ is within $\pm\varepsilon$ of the target;
* $w_{dt} \in \{0,1\}$ for $d \in C$ and $z_{dt} \in \{0,1\}$ for $d \in H$: $d$ is the cardiac
  or charge doctor on weekday $t$;
* $M^w, M^z, M^{wz} \ge 0$: the largest number of cardiac, charge and combined roles held by any
  one consultant in the week.

A consultant's points for the week are
$$
S_d = \sum_{t \in T} \sum_{k \in K_t} k \, x_{dtk} + a \, \bigl|\{t : d \in A_t\}\bigr|.
$$

## Constraints

Positions and fixed shifts:
$$
\begin{aligned}
&\textstyle\sum_{d \in W_t} x_{dtk} \le 1 && \forall t, k \in K_t \quad &&\text{(one consultant per position)}\\
&\textstyle\sum_{k \in K_t} x_{dtk} = 1 && \forall t \in T^w, d \in U_t &&\text{(every unassigned consultant gets a position)}\\
&x_{d, t, \pi_t(d)} = 1 && \forall t, d \text{ with a fixed shift} &&\text{(fixed shifts and requests)}
\end{aligned}
$$

Equity bands. With a big-$M$ constant $B$, for every $\varepsilon \in E$ and every $d$ with
$n_d > 0$:
$$
c - \varepsilon - B(1 - y^\varepsilon_d) \;\le\; \frac{S_d}{n_d} \;\le\; c + \varepsilon + B(1 - y^\varepsilon_d).
$$
If $y^\varepsilon_d = 1$, consultant $d$ is within $\varepsilon$ points per day of the target;
otherwise the constraint is relaxed. The bands are nested, so a consultant within 0.2 of the
target can set all three indicators to 1.

Roles, for every weekday $t$:
$$
\begin{aligned}
&\textstyle\sum_{d \in L_t \cap C} w_{dt} = 1 &&\text{(one cardiac doctor, from the call team)}\\
&\textstyle\sum_{d \in (L_t \cup U_t) \cap H} z_{dt} = 1 &&\text{(one charge doctor)}\\
&x_{d, t, \kappa_t} \ge z_{dt} && \forall d \in U_t \cap H \quad \text{(an unassigned charge doctor holds the charge position)}\\
&w_{dt} + z_{dt} \le 1 && \forall d \in C \cap H \quad \text{(not both roles on the same day)}\\
&z_{dt} + z_{d,t+1} \le 1 && \forall d \in H,\ t, t+1 \in T^w \quad \text{(no charge on consecutive days)}\\
&\textstyle\sum_{t} w_{dt} \le M^w,\ \sum_{t} z_{dt} \le M^z,\ \sum_{t} (w_{dt} + z_{dt}) \le M^{wz} && \text{(role spread)}
\end{aligned}
$$

## Objective

The model maximises, in order of priority,
$$
\max \;\; \underbrace{\sum_{\varepsilon \in E} \varepsilon \sum_{d \in D} y^\varepsilon_d}_{\text{equity}}
\;-\; \beta \, \underbrace{(M^w + M^z + M^{wz})}_{\text{role spread}}
\;+\; \gamma \, \underbrace{\sum_{t \in T^w} \sum_{d \in L_t \cap H} z_{dt}}_{\text{charge from call team}},
$$
with $\beta = 0.01$ and $\gamma = 0.001$. Equity dominates: a consultant within all three bands
contributes $1 + 0.5 + 0.2 = 1.7$. The small weights break ties in favour of spreading the
cardiac and charge roles, and then of giving the charge role to the call team, who stay late
anyway.

The target $c$ is itself a variable, so the model places it where as many consultants as
possible fall within the bands, close to the median of points per day worked.

## Solution

A typical weekly instance has about 1,200 variables, almost all binary, and 400 constraints. It
was solved with Gurobi [@gurobi] with a time limit of 60 seconds.
`r sum(apap$weeks$optimal)` of the `r n_weeks` weeks were solved to proven optimality; the rest
stopped at the time limit with a feasible solution.

# Results {#sec-results}

## Weekly objective

@fig-objectives shows the weekly target and the three objective terms. The target stays between
about 7 and 9 points per day worked. It dips in the first half of 2022 and rises afterwards.

```{r}
#| label: fig-objectives
#| fig-cap: "Weekly target $c$ and objective terms, one point per week."
#| fig-height: 4.5
plot_objectives(apap$weeks)
```

**TODO:** Explain the 2022 dip and the 2023 rise (staffing changes? new consultants?).

## Points per consultant

@fig-weekly-points compares raw weekly points with points per day worked. Raw points vary a lot,
mostly because consultants work different numbers of days in a week. Per day worked, the medians
are almost equal across consultants and across role groups.

```{r}
#| label: fig-weekly-points
#| fig-cap: "Weekly points per consultant (top) and points per day worked (bottom), coloured by the roles the consultant is qualified for."
#| fig-height: 6.5
plot_weekly_points(apap$points)
```

The long-run picture is in @fig-running-mean. After the first year, the cumulative points per day
worked settle into a band between
`r sprintf("%.1f", min(long_run$per_day))` and `r sprintf("%.1f", max(long_run$per_day))`.
A few consultants stay consistently above the rest.

```{r}
#| label: fig-running-mean
#| fig-cap: "Cumulative points per day worked for each consultant, from their 20th day worked."
#| fig-height: 3.5
plot_running_mean(apap$points)
```

The model does not carry points from one week to the next: every week is balanced on its own.
The long-run band in @fig-running-mean is therefore the result of many fair weeks, and a
consultant who is persistently a little above the target in each week is never compensated
later.

## Positions

@fig-position-heatmap shows how often each consultant held each position on weekdays. Most
consultants are spread across the whole range. The few exceptions have a narrow range of
positions, and they are consultants who are rarely or never in the call team. They therefore never get
the first position after a night on call, nor the last positions of the call team.

```{r}
#| label: fig-position-heatmap
#| fig-cap: "Share of weekdays each consultant held each peel-off position."
#| fig-height: 4.5
plot_position_heatmap(apap$assignments, apap$points)
```

## Holidays

Holiday shifts are fixed by the call schedule and are not decided by the model.
@fig-holidays shows that they are unevenly spread: some consultants worked the same holiday
many times.

```{r}
#| label: fig-holidays
#| fig-cap: "Holiday shifts per consultant and holiday."
#| fig-height: 3
plot_holidays(apap$assignments, apap$points)
```

# Discussion {#sec-discussion}

**Fairness within the week works.** Once the variation in days worked is taken into account, the
weekly model gives the consultants almost the same points per day worked, whatever roles they are
qualified for.

**Weekly equity does not imply long-run equity.** This is the main finding. APAP gives closely
balanced points per day worked within each week, but because successive weeks are optimised
independently, small systematic differences persist (@fig-running-mean). In the terms of
@sec-fairness, the weekly model is impartial but not necessarily unbiased. This result connects
the two strands motivating APAP: distributive equity within an allocation period does not
guarantee fairness across a sequence of periods. A sequence of locally fair allocations can
still produce a persistently uneven trajectory when past burden does not affect future priority.
Related distinctions arise in preference fulfilment on monthly duty rosters
[@gross2019satisfaction] and repetitive scheduling [@ajtai1998fairness;
@hermelin2025fairness].

**Carrying a balance between weeks.** The natural extension is to let each consultant enter a
week with a balance from the weeks before. With $P_d$ the points and $N_d$ the days worked by
consultant $d$ before the current week, and $\tau$ a long-run target, the balance is
$$
b_d = P_d - \tau N_d .
$$
A consultant with a positive balance has carried more than their share and enters the week with
a claim to earlier positions; the equity bands would then apply to cumulative instead of weekly
points per day worked. This is the claim and priority reasoning of @young1994equity applied over
time: past burden gives a claim, the claim gives priority, and the new allocation updates the
burden.

**Points are a proxy.** A position is not a time. Position 10 on a light day may mean going home
earlier than position 5 on a heavy day. Linking positions to actual release times would need
data on when consultants left, which is exactly the registry data that could not be obtained.
With such data, the points could be calibrated to hours, and the model evaluated against what
happened in practice.

**Fixed shifts dominate.** The call schedule, including holidays, fixes a large part of each
consultant's burden before the model runs. Holiday shifts in particular are unevenly shared
(@fig-holidays). Joint optimisation of the call schedule and the peel-off order is left for
future work.

**TODO:** Limitations of retrospective evaluation (see @sec-data); stakeholder feedback;
whether and how the model was used in practice.

## Related application

The same structure, a group that must stay until its part of a shared workload is done, appears
in grant review panels, where a member may leave a meeting once all the proposals they review have
been discussed. That application is being developed separately as a companion project.

# Conclusion {#sec-conclusion}

This paper formulates end-of-day anaesthesia relief as a repeated ordinal allocation problem.
APAP assigns a complete peel-off order for each day, jointly over a weekly planning horizon,
while respecting call-related positions, requests and clinical roles. It represents each
position as points and balances points per day worked, so that earlier and later release
positions are shared across consultants with different work patterns.

The retrospective application to 285 weeks of call schedules shows that the model can produce
closely balanced weekly allocations. It also exposes a limit of balancing each period on its
own: small differences can persist when past burden does not shape future allocations. Fairness
within a week is therefore distinct from fairness across the sequence of weeks. This empirical
result brings the two motivating perspectives together: equity theory helps define a fair
allocation within a period, while repeated-scheduling theory draws attention to how allocations
accumulate over time.

The model balances ordinal positions as a proxy for late-working burden; it does not measure
actual departure times. Future work could use departure-time data to calibrate the burden
associated with positions, and extend the model to carry each consultant's historical balance
between weeks. That would allow earlier allocations to inform later priorities and directly
address the gap between locally balanced weeks and long-run equity.

# References {.unnumbered}

::: {#refs}
:::

# Appendix: code and reproducibility {.appendix}

All code is in the project repository. From `code/`:

```bash
make all                       # schedules -> weekly instances -> solved weeks -> database
python -m data.apap_export     # database -> CSV files in data/results/
python -m data.build_docs_db   # CSV files -> compact SQLite database for the paper
```

The model is in `code/models/allocation_model.py`, the figures in `code/apap_figures.R`. The
paper's R code reads `docs/data/apap.sqlite` and is rendered with `quarto render` from the project
root; GitHub Actions installs the required R packages and runs the same render.
